---
title: "Continuous Data Example"
author: "Matt Wheeler Ph.D."
date: "12/05/2021"
# bibliography: "biblo.bib" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
#output:
#  html_document:
#    bibliography: "biblo.bib"  
#    toc: yes
#    toc_float: yes
#  word_document:
#    bibliography: "biblo.bib" 
#    toc: yes
#  pdf_document:
#    bibliography: "biblo.bib" 
#    toc: yes
---

```{r setup, include=FALSE}
library(bibtex)
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

## Basics of Toxic R Continuous run. 

This file shows ToxicR for dichotomous benchmark dose analyses using both single model and multiple model fits. The first thing we need to do is load the data.  From an excel file this can be done using the following R commands:


```{r load_data}
cont_data           <- matrix(0,nrow=5,ncol=4)
colnames(cont_data) <- c("Dose","Mean","N","SD")
cont_data[,1] <- c(0,50,100,200,400)
cont_data[,2] <- c(5.26,5.76,6.13,8.24,9.23)
cont_data[,3] <- c(20,20,20,20,20)
cont_data[,4]<-  c(2.23,1.47,2.47,2.24,1.56)
Y <- cont_data[,2:4]
```

### Hill fit
In what follows, I describe fitting the models using summary statistics.  One can also fit these models using the original data with no change to the function call except the Y is a $n\times 1$ now one column vector, and the dose vector is also $n\times 1$ and corresponds to each entry in the data vector. 

When summary data are used, ToxicR expects $Y$ to be a $n\times 3$ matrix. The first column of $Y$ is the the mean, the second colunn is the number of units on test, and the third column
is the observed standard deviation. In what follows, lets look at the Hill model fit, where the
Hill model is 
\begin{align}
f(dose) &= a + \frac{b\times dose}{c^d + [dose]^d}
\end{align}

```{r run_laplace_hill}
library(ToxicR)
library(ggplot2)
hill_fit <- single_continuous_fit(cont_data[,"Dose"],Y,
                                  model_type="hill")
```

### Object Extraction
The last line fits the Hill Bayesian Laplace model and puts all of the information one needs in the `hill_fit' object variable. This variable has the same structure as its dichotomous counterpart. By default the procedure assumes normal variance proportional to the mean. 
To see what model you fit, simply type the following line: 

```{r run_laplace_hill2}
hill_fit$full_model
hill_fit$prior
```

Here, we see the Hill model is fit using the Normal-NCV, or normal variance proportional 
to the mean. For the Hill model, you can also use 'normal' as a distribution option. 

```{r run_laplace_hill3, fig.height = 5, fig.width = 6}
hill_fit <- single_continuous_fit(cont_data[,"Dose"],
                                  cbind(cont_data[,"Mean"],cont_data[,"N"],cont_data[,"SD"]),
                                  model_type="hill",distribution = "normal",
                                  fit_type = "mcmc")
hill_fit$full_model
plot(hill_fit)

```

Notice how the `distribution' option controls the distribution. For the "hill", "power", and 
"polynomial" DR models you can choose either "normal" or "normal-ncv." or the "exp-3" and "exp-5" models you can also choose "lognormal."


```{r run_laplace_exp5, fig.height = 5, fig.width = 6}
exp5_fit <- single_continuous_fit(cont_data[,"Dose"],Y,
                                  model_type="exp-5",distribution = "lognormal",fit_type="laplace")
exp5_fit$full_model
plot(exp5_fit)
```

## Benchmark Dose Types

### Absolute Deviation
There are also other types of BMDs you can choose.  Here one can specify absolute deviation as
'abs,' which solves
\begin{align}
\mid f(\mbox{BMD}) - f(0) \mid = \mbox{BMR},
\end{align}
where $BMR$ is a specific cutoff value. In the example below, the abosolute difference
of  $\mbox{BMR} = 2.$

```{r run_laplace_hillad, fig.height = 5, fig.width = 6}
hill_sd_fit <- single_continuous_fit(cont_data[,"Dose"],Y,
                                  model_type="hill",distribution = "normal-ncv", fit_type="mcmc",
                                  BMD_TYPE="abs",BMR = 2)
hill_sd_fit$full_model
plot(hill_sd_fit)
```

### Standard Deviation

The standard deviation approach is the default approach, and this is the value that solves
\begin{align}
\mid f(\mbox{BMD}) - f(0) \mid = \mbox{BMR} \sigma_0,
\end{align}
and in this definition the $BMR \geq 0$ is the number of standard deviations the 
mean changes from no-exposure. In the example below,  the $\mbox{BMR} = 1.5.$

```{r run_laplace_hillsd, fig.height = 5, fig.width = 6}
hill_sd_fit <- single_continuous_fit(cont_data[,"Dose"],
                                  cbind(cont_data[,"Mean"],cont_data[,"N"],cont_data[,"SD"]),
                                  model_type="hill",distribution = "normal-ncv",fit_type="laplace",
                                  BMD_TYPE="sd",BMR = 1.5)
hill_sd_fit$full_model
plot(hill_sd_fit)
```


### Hybrid Deviation

The hybrid approach is a probabilistic approach that mimics risk for dichotomous 
data. Here, the BMD is the value that solves
\begin{align}
\frac{ Pr[y < Y_0 \mid \mbox{BMD}] - Pr[y > Y_0 \mid 0]}{1-Pr[y > Y_0 \mid 0]}  = \mbox{BMR} .
\end{align}
Here, $Pr[y > Y_0 \mid 0]$ is the probability that an 'adverse' response is 
observed at background dose, and  $0 < \mbox{BMR} < 1$ is the increase in probability
of seeing an adverse response at dose $\mbox{BMD}$. For this approach $Pr[y < Y_0 \mid 0]$
must be specified using the "point_p" option.  This option is only used when
"hybrid" is specified. The following shows a Hill fit, where the BMD with only a $2.5\%$ chance 
of being observed at no dose, but has a $12.25\%$ probability of being observed
at dose $\mbox{BMD}.$ Note: $\frac{0.1225 - 0.025}{0.975} = 0.1.$ For more information 
on this approach see Crump [@crump1995calculation]. 

```{r run_laplace_hillhybrid, fig.height = 5, fig.width = 6}
hill_hybrid_fit <- single_continuous_fit(cont_data[,"Dose"],
                                  cbind(cont_data[,"Mean"],cont_data[,"N"],cont_data[,"SD"]),
                                  model_type="hill",distribution = "normal-ncv",fit_type="mcmc",
                                  BMD_TYPE="hybrid",point_p = 0.025,BMR = 0.1)
hill_hybrid_fit$full_model
plot(hill_hybrid_fit)
```

### Relative Deviation

The relatived deviation approach [@slob2002dose] is the value that solves
\begin{align}
\mid f(\mbox{BMD}) - f(0) \mid = [1 + \mbox{BMR}] \times f(0),
\end{align}

where $\mbox{BMR} > 0$ is the value of increase/decrease from the background. 

```{r run_laplace_hillrd, fig.height = 5, fig.width = 6}
hill_rd_fit <- single_continuous_fit(cont_data[,"Dose"],
                                  cbind(cont_data[,"Mean"],cont_data[,"N"],cont_data[,"SD"]),
                                  model_type="hill",distribution = "normal-ncv", fit_type="mcmc",
                                  BMD_TYPE="rel",BMR = 0.1,samples = 50000)
hill_rd_fit$full_model
plot(hill_rd_fit)
```
Note: The above definitions are for for increasing data, `ToxicR' will determine
the direction of adversity and adjust computations accordingly. 

## Prior Specification

Like the dichotomous case you can modify the priors, to see what prior you want to modify
Again, this is easy we can type the following code to see what our current prior is: 

```{r run_laplace_cprior}
hill_sd_fit$prior

prior <- create_prior_list(normprior(0,1,-100,100),
                           normprior(0,1,-1e4,1e4),
                           lnormprior(0,1, 0, 100),
                           lnormprior(log(1),0.4215,0,18),
                           lnormprior(0,1,0,100),
                           normprior(0, 10,-100,100));

p_hill_ncv = create_continuous_prior(prior,"hill","normal-ncv")

prior <- create_prior_list(normprior(0,1,-100,100),
                           normprior(0,1,-1e4,1e4),
                           lnormprior(0,1, 0, 100),
                           lnormprior(log(1),0.4215,0,18),
                           normprior(0, 10,-100,100));

p_hill_norm = create_continuous_prior(prior,"hill","normal")
```

When specifying a prior, it is slightly different than the dichotomous case.  Here, 
we need to specify what distribution we are using too. This adds one or two parameters
depending on the parameters.  All models have a parameter called $\log(\sigma^2),$ which
is the variance term, and it is ALWAYS the last parameter in the model, and the  
prior is placed over the log of this value.  For the 'normal-NCV'
option, the variance is $\sigma^2 f(dose)^\rho,$ i.e., it has an additional parameter $\rho.$
The parameter $\rho$ must have a prior, which is $\mbox{log-Normal}(0,1)$ in the above
code. 

```{r run_laplace_cprior2, fig.height = 5, fig.width = 6}

hill_sd_a_fit <- single_continuous_fit(cont_data[,"Dose"],
                                  cbind(cont_data[,"Mean"],cont_data[,"N"],cont_data[,"SD"]),
                                  prior=p_hill_ncv,
                                  fit_type="laplace",
                                  BMD_TYPE="sd",BMR = 1.5)

hill_sd_b_fit <- single_continuous_fit(cont_data[,"Dose"],
                                  cbind(cont_data[,"Mean"],cont_data[,"N"],cont_data[,"SD"]),
                                  prior=p_hill_norm,
                                  fit_type="laplace",
                                  BMD_TYPE="sd",BMR = 1.5)

library(ggpubr)
figure <- ggarrange(plot(hill_sd_a_fit)+ggtitle(""), 
                    plot(hill_sd_b_fit)+ggtitle(""),
                    labels = c("Prior NCV", "Prior Normal"),
                    ncol = 1, nrow = 2)

figure
```

In the above code, we didn't need to specify the 'model_type' or 'distribution' because
this is done implicitly when you specify the prior. The above code works for the hill 
model, lets try it for a polynomial model. 


```{r run_laplace_polynomial, fig.height = 5, fig.width = 6}


poly_sd <- single_continuous_fit(cont_data[,"Dose"],Y,
                                  distribution="normal",model_type="polynomial",
                                  degree = 4, 
                                  fit_type="laplace",
                                  BMD_TYPE="sd",BMR = 0.5)

plot(poly_sd)

```

## Model Averaging

Beyond single model fits, one can also use model averging for continuous models. 
As a default 10 models are fit. Here, the 'hill' and 'power' options are fit under
the 'normal' and 'normal-ncv' distribution assumptions.  Additionally to the 
'normal' and 'normal-ncv' distributions the 'exp-3' and 'exp-5' also use the 
'lognormal' distribution. Note that the polynomial model is not used because
of constraint issues. 

```{r run_laplace_MA, fig.height = 5, fig.width = 6}
ma_sd_mcmc <- ma_continuous_fit(cont_data[,"Dose"],Y, fit_type="mcmc",
                             BMD_TYPE="sd",BMR = 0.5,samples = 50000)

ma_sd_laplace <- ma_continuous_fit(cont_data[,"Dose"],Y, fit_type="laplace",
                                   BMD_TYPE="sd",BMR = 0.5)

plot(ma_sd_mcmc)
cleveland_plot(ma_sd_laplace)
MAdensity_plot(ma_sd_mcmc)
```

Here, 'cleveland_plot' and 'MAdensity_plot' can be used in the same way as in 
dichotomous plot. Like the dichotomous case, we can create a prior list for our model
average. 

```{r run_laplace_MA_2}
prior <- create_prior_list(normprior(0,1,-100,100),
                           normprior(0,1,-1e4,1e4),
                           lnormprior(0,1, 0, 100),
                           lnormprior(log(1),0.4215,0,18),
                           lnormprior(0,1,0,100),
                           normprior(0, 10,-100,100));

p_hill_ncv = create_continuous_prior(prior,"hill","normal-ncv")

prior <- create_prior_list(normprior(0,1,-100,100),
                           normprior(0,1,-1e4,1e4),
                           lnormprior(0,1, 0, 100),
                           lnormprior(log(1),0.4215,0,18),
                           normprior(0, 10,-100,100));

p_hill_norm = create_continuous_prior(prior,"hill","normal")

prior <- create_prior_list(normprior(0,1,-100,100),
                           normprior(0,1,-1e4,1e4),
                           lnormprior(0,1, 0, 100),
                           lnormprior(log(1),0.4215,0,18),
                           normprior(0, 10,-100,100));

p_exp5_norm = create_continuous_prior(prior,"exp-5","normal")

prior <- create_prior_list(normprior(0,1,-100,100),
                           normprior(0,1,-1e4,1e4),
                           lnormprior(log(1),0.4215,0,18),
                           normprior(0, 10,-100,100));
p_power_norm = create_continuous_prior(prior,"power","normal")

prior <- create_prior_list(normprior(0,1,-100,100),
                           normprior(0,1,-1e4,1e4),
                           lnormprior(log(1),0.4215,0,18),
                           normprior(0, 10,-100,100));
p_exp3_norm = create_continuous_prior(prior,"exp-3","normal")
prior_list = list(p_exp3_norm,p_hill_norm,p_exp5_norm,p_power_norm)
```

With a list of priors one can then run a model average based upon 
a user specified model space.   Here, all of the same benchmark dose options 
of 'single_continuous_fit' apply.

```{r run_laplace_MA_3, fig.height = 5, fig.width = 6}
ma_sd_mcmc_2 <- ma_continuous_fit(cont_data[,"Dose"],Y, fit_type= "laplace",
                             BMD_TYPE="sd",BMR = 0.5,samples = 50000,model_list=prior_list)
plot(ma_sd_mcmc_2)
```


## References

Wout Slob, Dose-Response Modeling of Continuous Endpoints, Toxicological Sciences, Volume 66, Issue 2, April 2002, Pages 298–312

Crump, Kenny S., Calculation of benchmark doses from continuous data, Risk Analysis, Volume 15, Issue 1, February 1995, Pages 79-89
